Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-10-18 VMCAI-revised-version/final.tex
ideally effective vs ideally-effective, kept only the one with '-'

To obtain decidability results, we must introduce some notions of \emph{effectiveness}. We use a mixture of the notions defined both by Blondin and al. in \cite{DBLP:journals/lmcs/BlondinFM17}) and by Halfon in \cite{DBLP:phd/hal/Halfon18}.
%			 that will be essential for solving various resilience problems in different models of WSTS. 
First, to simplify, we suppose that all classes of considered ordered transition systems $\mathscr{S}=(S, \rightarrow, \leq)$ satisfy the five following properties:
\begin{enumerate}
\item there exists an algorithm that decides whether $s \rightarrow t$ is true or not, for any states $s,t \in S$,
%
\item there is a computational representation for $S$ for which membership in $S$ is decidable,
%
%			We also suppose that all considered wqos $(S,\leq)$ satisfy the three following conditions : 
\item $\leq$ is decidable,
\item there is a computational representation for $Ideals(S)$ for which membership in $Ideals(S)$ is decidable, and
\item  inclusion of ideals is decidable. 
\end{enumerate}
%
Blondin and al. showed (Lemma 4.3 in \cite{DBLP:journals/lmcs/BlondinFM17}) that under the previous hypotheses, one are able to enumerate \emph{downward closed sets} (by their finite decomposition in ideals), to decide inclusion between downward closed sets and to decide if a state belongs to a given finite set of ideals. But the five properties are not sufficient to compute the complementary and the intersection of downward closed sets.
%
Halfon introduced \emph{ideally effective wqo} as wqo that essentially allow to compute representations of (principal) ideals $\downarrow s$, of the complementary of an ideal, of the complementary of filters (a filter is a set $\uparrow s$ for $s \in S$) and to compute representations of finite intersections of filters and finite intersections of ideals. Most well-known wqo are ideally effective . \cite{DBLP:phd/hal/Halfon18}.
%	 (hence of downward closed sets).

%
Starting now, we assume that all considered WBTS $\mathscr{S}=(S, \rightarrow, \leq)$ satisfy the five previous properties and that the wqo $\leq$ is ideally effective. We refer to such WBTS as \emph{effective}.

@@

\begin{proof}
The resilience problem can be reformulated as 
$\Bad \subseteq  \pred^*(\Safe)$ that is equivalent to $\Bad \cap (S \setminus \pred^*(\Safe)) = \emptyset$.
The set $\pred^*(\Safe)$ is upward-closed, since $\Safe$ is upward-closed. Since $\mathscr{S}=(S,\rightarrow, \leq)$ is $\pred$-basis-effective, we can compute a basis of $\pred^*(\Safe)$.
Since the wqo is ideally effective, we can compute the intersection of
$\Bad$
and
$S \setminus \pred^*(\Safe)$,
which are both downward-closed, and test if this intersection is empty or not. Hence, the resilience problem is decidable.
%%%%%%%
\alain{faire les deux autres résilience}
\end{proof} \qed
%				The completion-$\post$-effective hypothesis is needed to decide {\sc Resilience}.
%%%
%%%
%Indeed, consider the family $\{ f_j: \N^2 \to \N^2\}$ of 
%increasing recursive functions from \cite{FMP-ic04} defined as
%
%$ f_j(n,k) = \begin{dcases*} (n,0) & \text{ if } k=0 \text{ and } TM$_j$ \text{ runs for more than } n \text{ steps} \\
%		(n,n+k) &\text{ otherwise,} \end{dcases*} $
%		
%	\noindent	
%	where TM$_j$ is the $j$-th Turing machine (in a classical enumeration)
%	which moreover begins
%by writing the integer $j$ on its tape,	
%	and consider additionnally the function $g : \N^2 \to \N^2$ defined by $g(n,k) = (n+1,k)$. 
%	The
%	transition system 
%	$S_j = (\N^2, \{f_j, g\}, \leq)$ 
%	is a WSTS
%	and
%	has the property that
%	it is
%	$\uparrow (1,1)$-resilient
%	iff TM$_j$ halts on input $j$,
%	since $(1,1)$ is coverable from $(0,0)$
%	iff TM$_j$ halts on input $j$. 
%	Hence there is no Turing machine which correctly determines 
%	resilience and halts whenever its input is a
%	WSTS.
%Hence resilience for WSTS in general is undecidable.

@@

\begin{proof}
\alain{preuve pas claire, à réécrire}
% \alain{les reset-VASS sont-ils effective $\post^*$-basis ?}
% \mathieu{non, sinon on aurait décidabilité pour la state résilience}
%This stems from the fact that it is undecidable for Minsky machines with more than one counter,
%which can be simulated by WSTS.
First, let us show that {\sc Resilience} is undecidable for Minsky machine with at least $3$ counters.
%	This is due to the undecidability of $2$-counter Minsky machine termination~\cite{Min61,Min67}.
From a $2$-counter Minsky machine $M$ with final control-state $q_f \in Q_M$, one can construct a $3$-counter Minsky machine $M'$ with final control-state $q_f' \in Q_{M'}$
such that machine $M$ terminates for all inputs iff machine $M'$ can reach $q_f'(0,0,0)$ from any input in $Q_{M'} \times \uparrow (0,0,1)$. 
%			with at least $1$ on its third counter. 
We build $M'$ to simulate $M$ until it reaches $q_f$, then decrease the first two counters until they reach $0$, then, and only then, finally decrease the third counter until it reaches $0$.
Remark that in the construction, from any input in $Q_{M'} \times \uparrow(0,0,1)$, it is not possible to reach an element in $Q_{M'} \times \{ (1,0,0), (0,1,0),(1,1,0)\}$.
Based on this construction, the problem of deciding whether the downward closed set $Q_{M'} \times \downarrow (1,1,0)$ is reachable in the $3$-counter Minsky machine $M'$
from any input in $Q_{M'} \times \uparrow(0,0,1)$ is equivalent to the termination of machine $M$, that is undecidable ~\cite{Min61}. 
Hence {\sc Resilience} is undecidable for Minsky machines.
\alain{écrire qqchose ici}
Now, executions of Minsky machines can be simulated by reset-VASS~\cite{araki1976PN}~\cite{dufourd1998reset}. 
% \alain{\cite{araki1976PN} simule-t-il les Minsky machines et comment ?}
%Reset-VASS extend the basic VASS model with special “reset
%transitions” that set to $0$ some coordinates in the vector. 
%			Let us recall their definition here.
%      	it is well known that reset-VASS are WSTS~\cite{dufourd1998reset}.
The simulation operates by building a reset-VASS in which
there exists
 a sequence from 
an initial state %$(s,x_1 \ldots x_n)$
to
a state of the form $ q(0,  \ldots 0)$ if and only if
there exists
a sequence
 in the Minsky machine
 from a corresponding initial state
 to
 a state of the form $q_f(0,0,0)$
  (from Theorem~$5$ in \cite{araki1976PN}).
Hence $Q_{M'} \times \downarrow (1,1,0)$-resilience hold in the Minsky machine $M'$ iff 
resilience hold in the reset-VASS simulating $M'$.
Since reset-VASS can simulate executions of a Minsky machine, {\sc Resilience} is undecidable for reset-VASS and hence for  $\pred$-basis-effective, 
$\omega^2$-WSTS with strong compatibility as well.\qed
%\alain{déjà dit avant en mieux omega2 et strong comp et $\post$-effective mais reset-VASS apportent effective $\pred$-basis et ideally effective....}
\end{proof}

@@

\begin{theorem}\label{post srp}
{\sc State-resilience} is decidable for $\post^*$-basis-effective WSTS
when
$\Safe=\uparrow \Safe$. 
\end{theorem}

>>>>

To obtain decidability results, we must introduce some notions of \emph{effectiveness}. We use a mixture of the notions defined both by Blondin and al. in \cite{DBLP:journals/lmcs/BlondinFM17}) and by Halfon in \cite{DBLP:phd/hal/Halfon18}.
%			 that will be essential for solving various resilience problems in different models of WSTS. 
First, to simplify, we suppose that all classes of considered ordered transition systems $\mathscr{S}=(S, \rightarrow, \leq)$ satisfy the five following properties:
\begin{enumerate}
\item there exists an algorithm that decides whether $s \rightarrow t$ is true or not, for any states $s,t \in S$,
%
\item there is a computational representation for $S$ for which membership in $S$ is decidable,
%
%			We also suppose that all considered wqos $(S,\leq)$ satisfy the three following conditions : 
\item $\leq$ is decidable,
\item there is a computational representation for $Ideals(S)$ for which membership in $Ideals(S)$ is decidable, and
\item  inclusion of ideals is decidable. 
\end{enumerate}
%
Blondin and al. showed (Lemma 4.3 in \cite{DBLP:journals/lmcs/BlondinFM17}) that under the previous hypotheses, one are able to enumerate \emph{downward closed sets} (by their finite decomposition in ideals), to decide inclusion between downward closed sets and to decide if a state belongs to a given finite set of ideals. But the five properties are not sufficient to compute the complementary and the intersection of downward closed sets.
%
Halfon introduced \emph{ideally-effective wqo} as wqo that essentially allow to compute representations of (principal) ideals $\downarrow s$, of the complementary of an ideal, of the complementary of filters (a filter is a set $\uparrow s$ for $s \in S$) and to compute representations of finite intersections of filters and finite intersections of ideals. Most well-known wqo are ideally-effective . \cite{DBLP:phd/hal/Halfon18}.
%	 (hence of downward closed sets).

%
Starting now, we assume that all considered WBTS $\mathscr{S}=(S, \rightarrow, \leq)$ satisfy the five previous properties and that the wqo $\leq$ is ideally-effective. We refer to such WBTS as \emph{effective}.

@@

\begin{proof}
The resilience problem can be reformulated as 
$\Bad \subseteq  \pred^*(\Safe)$ that is equivalent to $\Bad \cap (S \setminus \pred^*(\Safe)) = \emptyset$.
The set $\pred^*(\Safe)$ is upward-closed, since $\Safe$ is upward-closed. Since $\mathscr{S}=(S,\rightarrow, \leq)$ is $\pred$-basis-effective, we can compute a basis of $\pred^*(\Safe)$.
Since the wqo is ideally-effective, we can compute the intersection of
$\Bad$
and
$S \setminus \pred^*(\Safe)$,
which are both downward-closed, and test if this intersection is empty or not. Hence, the resilience problem is decidable.
%%%%%%%
\alain{faire les deux autres résilience}
\end{proof} \qed
%				The completion-$\post$-effective hypothesis is needed to decide {\sc Resilience}.
%%%
%%%
%Indeed, consider the family $\{ f_j: \N^2 \to \N^2\}$ of 
%increasing recursive functions from \cite{FMP-ic04} defined as
%
%$ f_j(n,k) = \begin{dcases*} (n,0) & \text{ if } k=0 \text{ and } TM$_j$ \text{ runs for more than } n \text{ steps} \\
%		(n,n+k) &\text{ otherwise,} \end{dcases*} $
%		
%	\noindent	
%	where TM$_j$ is the $j$-th Turing machine (in a classical enumeration)
%	which moreover begins
%by writing the integer $j$ on its tape,	
%	and consider additionnally the function $g : \N^2 \to \N^2$ defined by $g(n,k) = (n+1,k)$. 
%	The
%	transition system 
%	$S_j = (\N^2, \{f_j, g\}, \leq)$ 
%	is a WSTS
%	and
%	has the property that
%	it is
%	$\uparrow (1,1)$-resilient
%	iff TM$_j$ halts on input $j$,
%	since $(1,1)$ is coverable from $(0,0)$
%	iff TM$_j$ halts on input $j$. 
%	Hence there is no Turing machine which correctly determines 
%	resilience and halts whenever its input is a
%	WSTS.
%Hence resilience for WSTS in general is undecidable.

@@

\begin{proof}
\alain{preuve pas claire, à réécrire}
% \alain{les reset-VASS sont-ils effective $\post^*$-basis ?}
% \mathieu{non, sinon on aurait décidabilité pour la state résilience}
%This stems from the fact that it is undecidable for Minsky machines with more than one counter,
%which can be simulated by WSTS.
First, let us show that {\sc Resilience} is undecidable for Minsky machine with at least $3$ counters.
%	This is due to the undecidability of $2$-counter Minsky machine termination~\cite{Min61,Min67}.
From a $2$-counter Minsky machine $M$ with final control-state $q_f \in Q_M$, one can construct a $3$-counter Minsky machine $M'$ with final control-state $q_f' \in Q_{M'}$
such that machine $M$ terminates for all inputs iff machine $M'$ can reach $q_f'(0,0,0)$ from any input in $Q_{M'} \times \uparrow (0,0,1)$. 
%			with at least $1$ on its third counter. 
We build $M'$ to simulate $M$ until it reaches $q_f$, then decrease the first two counters until they reach $0$, then, and only then, finally decrease the third counter until it reaches $0$.
Remark that in the construction, from any input in $Q_{M'} \times \uparrow(0,0,1)$, it is not possible to reach an element in $Q_{M'} \times \{ (1,0,0), (0,1,0),(1,1,0)\}$.
Based on this construction, the problem of deciding whether the downward closed set $Q_{M'} \times \downarrow (1,1,0)$ is reachable in the $3$-counter Minsky machine $M'$
from any input in $Q_{M'} \times \uparrow(0,0,1)$ is equivalent to the termination of machine $M$, that is undecidable ~\cite{Min61}. 
Hence {\sc Resilience} is undecidable for Minsky machines.
\alain{écrire qqchose ici}
Now, executions of Minsky machines can be simulated by reset-VASS~\cite{araki1976PN}~\cite{dufourd1998reset}. 
% \alain{\cite{araki1976PN} simule-t-il les Minsky machines et comment ?}
%Reset-VASS extend the basic VASS model with special “reset
%transitions” that set to $0$ some coordinates in the vector. 
%			Let us recall their definition here.
%      	it is well known that reset-VASS are WSTS~\cite{dufourd1998reset}.
The simulation operates by building a reset-VASS in which
there exists
 a sequence from 
an initial state %$(s,x_1 \ldots x_n)$
to
a state of the form $ q(0,  \ldots 0)$ if and only if
there exists
a sequence
 in the Minsky machine
 from a corresponding initial state
 to
 a state of the form $q_f(0,0,0)$
  (from Theorem~$5$ in \cite{araki1976PN}).
Hence $Q_{M'} \times \downarrow (1,1,0)$-resilience hold in the Minsky machine $M'$ iff 
resilience hold in the reset-VASS simulating $M'$.
Since reset-VASS can simulate executions of a Minsky machine, {\sc Resilience} is undecidable for reset-VASS and hence for  $\pred$-basis-effective, 
$\omega^2$-WSTS with strong compatibility as well.\qed
%\alain{déjà dit avant en mieux omega2 et strong comp et $\post$-effective mais reset-VASS apportent effective $\pred$-basis et ideally-effective....}
\end{proof}

@@

\begin{theorem}\label{post srp}
{\sc State-resilience} is decidable for $\pred$-effective, $\post^*$-basis-effective WSTS
when
$\Safe=\uparrow \Safe$. 
\end{theorem}
